---
title: "Installing and Managing _Bioconductor_ Packages"
author:
- name: Marcel Ramos
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
- name: Martin Morgan
  affiliation: Roswell Park Comprehensive Cancer Center, Buffalo, NY
output:
  BiocStyle::html_document:
      toc: true
vignette: |
  %\VignetteIndexEntry{Installing and Managing Bioconductor Packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

Use the [BiocManager][1] package to install and manage packages from the
_[Bioconductor][2]_ project for the statistical analysis and comprehension of
high-throughput genomic data.

Current _Bioconductor_ packages are available on a 'release' version intended
for every-day use, and a 'devel' version where new features are introduced. A
new release version is created every six months. Using the [BiocManager][1]
package helps users install packages from the same release.

# Basic use

Use standard _R_ installation procedures to install the
[BiocManager][1] package. This command is requried only once per _R_
installation.

```{r, eval = FALSE}
chooseCRANmirror()
install.packages("BiocManager")
```

Install _Bioconductor_ (or CRAN) packages with

```{r, eval = FALSE}
BiocManager::install(c("GenomicRanges", "Organism.dplyr"))
```

Installed packages can be updated to their current version with

```{r, eval = FALSE}
BiocManager::install()
```

Use `version()` to discover the version of _Bioconductor_ currently in
use.

```{r}
BiocManager::version()
```

_Bioconductor_ packages work best when they are all from the same release. Use
`valid()` to identify packages that are out-of-date or from unexpected
versions.

```{r}
BiocManager::valid()
```

`valid()` returns an object that can be queried for detailed
information about invalid packages, as illustrated in the following
screen capture

```
> v <- valid()
Warning message:
6 packages out-of-date; 0 packages too new
> names(v)
[1] "out_of_date" "too_new"
> head(v$out_of_date, 2)
    Package LibPath
bit "bit"   "/home/mtmorgan/R/x86_64-pc-linux-gnu-library/3.5-Bioc-3.8"
ff  "ff"    "/home/mtmorgan/R/x86_64-pc-linux-gnu-library/3.5-Bioc-3.8"
    Installed Built   ReposVer Repository
bit "1.1-12"  "3.5.0" "1.1-13" "https://cran.rstudio.com/src/contrib"
ff  "2.2-13"  "3.5.0" "2.2-14" "https://cran.rstudio.com/src/contrib"
>
```

Packages available for your version of _Bioconductor_ can be
discovered with `available()`; the first argument can be used to
filter package names based on a regular expression, e.g., 'BSgenome' package available for _Homo sapiens_
```{r}
avail <- BiocManager::available()
length(avail)
BiocManager::available("BSgenome.Hsapiens")
```

Questions about installing and managing _Bioconductor_ packages should be
addressed to the [_Bioconductor_ support site][3].

# Advanced use

## Changing version

Use the `version=` argument to update all packages to a specific _Bioconductor_
version

```{r, eval = FALSE}
BiocManager::install(version="3.7")
```

_Bioconductor_ versions are associated with specific _R_ versions, as
summarized [here][5]. To use the most recent version of _Bioconductor_
may require installing a new version of _R_.

A special version, `version="devel"`, allows use of _Bioconductor_
packages that are under development.

## Managing multiple versions

It is possible to have multiple versions of _Bioconductor_ installed on the
same computer. A best practice is to [create an initial _R_ installation][6].
Then create and use a library for each version of _Bioconductor_. The library
will contain all _Bioconductor_, CRAN, and other packages for that version of
_Bioconductor_. We illustrate the process assuming use of _Bioconductor_
version 3.7, available using _R_ version 3.5

Create a directory to contain the library (replace `USER_NAME` with your user
name on Windows)

- Linux: `~/R/3.5-Bioc-3.7`
- macOS: `~/Library/R/3.5-Bioc-3.7/library`
- Windows: `C:\Users\USER_NAME\Documents\R\3.5-Bioc-3.7`

Set the environment variable `R_LIBS_USER` to this directory, and invoke _R_.
Command line examples for Linux are

- Linux: `R_LIBS_USER=~/R/3.5-Bioc-3.7 R`
- macOS: `R_LIBS_USER=~~/Library/R/3.5-Bioc-3.7/library R`
- Windows: `cmd /C "set R_LIBS_USER=C:\Users\USER_NAME\Documents\R\3.5-Bioc-3.7 && R"`

Once in _R_, confirm that the version-specific library path has been set

```{r, eval = FALSE}
.libPaths()
```

On Linux and macOS, create a bash alias to save typing, e.g.,

- Linux: `alias Bioc3.7='R_LIBS_USER=~/R/3.5-Bioc-3.7 R'`
- macOS: `alias Bioc3.7='R_LIBS_USER=~/Library/R/3.5-Bioc-3.7/library R'`

Invoke these from the command line as `Bioc3.7`.

On Windows, create a shortcut. Go to My Computer and navigate to a directory
that is in your PATH. Then right-click and choose New->Shortcut.
In the "type the location of the item" box, put:

```
cmd /C "set R_LIBS_USER=C:\Users\USER_NAME\Documents\R\3.5-Bioc-3.7 && R"
```

Click "Next". In the "Type a name for this shortcut" box, type `Bioc-3.7`.

[1]: https://cran.r-project.org/package=BiocManager
[2]: https://bioconductor.org
[3]: https://support.bioconductor.org
[5]: https://bioconductor.org/about/release-announcements/
[6]: https://cran.R-project.org/
